---
title: "exercise 11"
format: html
editor: visual
---

## Exercise 11

```{r setup}
install.packages("naniar")
library(naniar)
library(tidyverse)
library(ggplot2)
library(MASS)
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/main/Mammal_lifehistories_v2.txt"
d <- read_tsv(f, col_names = TRUE)
```

1.  Replace -999 with NA

```{r}
d <- d |> replace_with_na_all(condition = ~.x == -999)
names(d)
```

2.  Data Cleaning

    Drop the variables **litter size** and **refs**.

    Rename the variable **max. life(mo)** to **maxlife(mo)**.

    Rename the variable **wean mass(g)** to **weanmass(g)**.

```{r}
#i know i could do this all at once but i don't wanna
d <- d |> select(-c(`litter size`, refs))
d <- d |> mutate( `maxlife(mo)` = `max. life(mo)`)
d <- d |> mutate(`weanmass(g)` = `wean mass(g)`)
d <- d |> select(- "wean mass(g)") |> select(- "max. life(mo)")
names(d)
#for some reasons the select function is not working 
```

3.  Log transform all of the other numeric variables

```{r}
d <- d |> mutate(`mass(g)` = log(`mass(g)`), `gestation(mo)` = log(`gestation(mo)`), `newborn(g)` = log(`newborn(g)`), `weaning(mo)` = log(`weaning(mo)`), `AFR(mo)` = log(`AFR(mo)`), `litters/year` = log(`litters/year`), `maxlife(mo)` = log(`maxlife(mo)`), `weanmass(g)` = log(`weanmass(g)`))
```

4.  Regress the (now log transformed) *age* \[**gestation(mo)**, **weaning(mo)**, **AFR(mo)** (i.e., age at first reproduction), and **maxlife(mo)** (i.e., maximum lifespan)\] and *mass* \[**newborn(g)** and **weanmass(g)**\] variables on (now log transformed) overall body **mass(g)** and add the residuals to the dataframe as new variables \[**relGest**, **relWean**, **relAFR**, **relLife**, **relNewbornMass**, and **relWeaningMass**\].

```{r}
#Age Regressions
lm1 <- lm(`mass(g)` ~ `gestation(mo)`, data = d, na.action=na.exclude)
lm2 <- lm(`mass(g)` ~ `weaning(mo)`, data = d, na.action=na.exclude)
lm3 <- lm(`mass(g)` ~ `AFR(mo)`, data = d, na.action=na.exclude)

#Lifespan Regression
lm4 <- lm(`mass(g)` ~ `maxlife(mo)`, data = d, na.action=na.exclude)

# Mass Regression
lm5 <- lm(`mass(g)` ~ `newborn(g)`, data = d, na.action=na.exclude)
lm6 <- lm(`mass(g)` ~ `weanmass(g)`, data = d, na.action=na.exclude)

d <- d |> mutate(relGest = resid(lm1), relWean = resid(lm2), relAFR = resid(lm3), relLife = resid(lm4), relNewbornMass = resid(lm5), relWeaningMass = resid(lm6))
names(d)
```

5.  Plot residuals of max lifespan (**relLife**) in relation to **Order**. Which mammalian orders have the highest residual lifespan?

```{r}
#| warning: false
#sorry this is an ugly plot
p1 <- ggplot(d, aes(x = order, y = relLife)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
p1
```

*The highest residual maximum life spans are in the orders Artiodactyla (the even-toed ungulates), Cetacea (whales and dolphins), and Perissodactyla (the odd-toed ungulates). The mean residual life span for these orders is above 1.5 according to my plot.*

Plot residuals of newborn mass (**relNewbornMass**) in relation to **Order**. Which mammalian orders have the have highest residual newborn mass?

```{r}
#| warning: false
p2 <- ggplot(d, aes(x = order, y = relNewbornMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
p2
```

*The highest mean residual masses at birth belong to the orders Carnivora, Perissodactyla, Probiscidea (elephants), and Tubulidentata (aardvark).*

Plot residuals of weaning mass (**relWeaningMass**) in relation to **Order**. Which mammalian orders have the have highest residual weaning mass?

```{r}
#| warning: false
p3 <- ggplot(d, aes(x = order, y = relWeaningMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
p3
```

The highest mean residial masses at weaning belong to Xenarthra (sloths), Tubulidentata, Hyracoidea (hyraxes), and Sirenia (manatees and dugongs). Interestingly, all of these orders have very few representative species.

6.  Run models and a model selection process to evaluate what (now log transformed) variables best predict each of the two response variables, **maxlife(mo)** and **AFR(mo)**, from the set of the following predictors: **gestation(mo)**, **newborn(g)**, **weaning(mo)**, **weanmass(g)**, **litters/year**, and overall body **mass(g)**.

*For maxlife(mo) I will use a forward approach and for AFR(mo) I will use a backward approach, both using the F statistic to determine the best model.*

```{r}
#| warning: false

#first, drop rows with NA values for any of the predictors or outcomes 
#yes this could be cleaner
nrow(d) #1440
d <- d |> drop_na(`maxlife(mo)`) |> drop_na(`AFR(mo)`) |> drop_na(`gestation(mo)`) |> drop_na(`newborn(g)`) |> drop_na(`weaning(mo)`) |> drop_na(`weanmass(g)`) |> drop_na(`litters/year`) |> drop_na(`mass(g)`)
nrow(d) #206

#m0 reflects the maxlife at the intercept, without any predictors
m0 <- lm(data = d, `maxlife(mo)` ~ 1)
add1(m0, scope = .~. + `gestation(mo)` + `newborn(g)` + `weanmass(g)` + `litters/year` + `mass(g)`, test = "F")
#for all variables that have a p < 0.05, adding that variable ALONE would add explanatory power to our null model
```

```{r}
#| warning: false
#add the predictor with the highest F statistic to the null model
#add gestation length to the model

m1 <- update(m0, formula = .~. + `gestation(mo)`)
#check if the remaining predictors are still significantly contributing to m1
add1(m1, scope = .~. + `newborn(g)` + `weanmass(g)` + `litters/year` + `mass(g)`, test = "F")
#all predictors are still significant so I can't immediately count any out
```

```{r}
#| warning: false
#add litters/year to the model
m2 <- update(m1, formula = .~. + `litters/year`)
add1(m2, scope = .~. + `newborn(g)` + `weanmass(g)` + `mass(g)`, test = "F")
#all are still significant
```

```{r}
#| warning: false
#adding mass(g)
m3 <- update(m2, formula = .~. + `mass(g)`)
add1(m3, scope = .~. + `newborn(g)` + `weanmass(g)`, test = "F")
#neither newborn mass or wean mass are signfiicant anymore, so m3 is my final model
```

```{r}
#AFR(mo) model
m.full <- lm(data = d, `AFR(mo)` ~  `gestation(mo)` + `newborn(g)` + `weanmass(g)` + `litters/year` + `mass(g)`)
drop1(m.full, test = "F")
#DROP the predictor with the lowest F statistic
#in this case, weanmass(g) where F = 0.1094
```

```{r}
m1 <- update(m.full, formula = .~. - `weanmass(g)`)
drop1(m1, test = "F")
#drop newborn(g), F = 0.2972
```

```{r}
m2 <- update(m1, formula = .~. - `newborn(g)`)
drop1(m2, test = "F")
#drop mass(g) where F = 1.3875
```

```{r}
m3 <- update(m2, formula = .~. - `mass(g)`)
drop1(m3, test = "F")
#both the remaining predictors are significant so this is my final model
```

-   For each of the two response variables, indicate what is the best model overall based on AICc and how many models have a delta AICc of 4 or less?

*Again, I will begin with the full model for AFR and use an additive approach for max lifespan.*

```{r}
s_AFR <- stepAIC(m.full, scope = .~., direction = "both")
#wean mass, mass, and newborn mass all reduce the AIC when included in the model
#the best model for AFR would be including mass, weanmass, and newborn mass but no other variables
```

*For predicting AFR, weanmass(g), mass(g), and newborn(g) all reduce the AIC when included in the full model. The best model is one that takes gestation(mo) and litters/year as predictor variables, and does not include the mass variables.*

```{r}
s <- stepAIC(m0,
scope = .~. + `gestation(mo)` + `newborn(g)` + `weanmass(g)` + `litters/year` + `mass(g)`, direction = "both")
#the AIC is most reduced by gestation(mo), so it gets added to the model
#the models have predictors added until any further predictors would no longer reduce the model
```

*The best model for maxlife(mo) is one that includes newborn(g), mass(g), litters/year, and gestation(mo), but does not include weanmass(g).*

*None of the models have an AIC \<= 4.*

-   What variables, if any, appear in all of this set of “top” models?

*For both maxlife and AFR, gestation length and litters/year are signfiicant predictors.*

-   Calculate and plot the model-averaged coefficients and their CIs across this set of top models.

```{r}
library(MuMIn)

##Max Lifespan
m.life <- lm(data = d, `maxlife(mo)` ~  `gestation(mo)` + `newborn(g)` + `weanmass(g)` + `litters/year` + `mass(g)`, na.action = na.fail)
mods.life <- dredge(m.life)

mavg.life <- summary(model.avg(mods.life, subset = delta <=4, fit = TRUE))
mavg.life$msTable

confint(mavg.life)
```

```{r}
##AFR
m.afr <- lm(data = d, `AFR(mo)` ~  `gestation(mo)` + `newborn(g)` + `weanmass(g)` + `litters/year` + `mass(g)`, na.action = na.fail)
mods.afr <- dredge(m.afr)

mavg.afr <- summary(model.avg(mods.afr, subset = delta <=4, fit = TRUE))
mavg.afr$msTable

confint(mavg.afr)
```
